---
title: "P8105-HW3-lj2607"
author: "Li Jiang"
date: "2024-10-11"
output: github_document
---

```{r}
#load the required packages
library(tidyverse)
library(lubridate)
library(readxl)
library(stringr)
```

#Problem1

```{r}
#load the required dataset
library(p8105.datasets)
data("ny_noaa")

#view the dataset
head(ny_noaa)
```

```{r}
#Data cleaning
ny_noaa_df = ny_noaa %>%
  janitor::clean_names()%>%
  drop_na() %>%
  mutate(
    year = year(date), 
    month = month(date),
    day = day(date),
    tmin = as.numeric(tmin) / 10,
    tmax = as.numeric(tmax) / 10,
    prcp = as.numeric(prcp),
    snowfall = as.numeric(snow)
)

#look at the data
view(ny_noaa_df)
```

```{r}
#Checking the most common value in snowfall
snowfall_distribution_plot <- ny_noaa |>
  filter(snow > 0 & snow < 100) |>
  ggplot(aes(x = factor(year), y = snow)) +
  geom_boxplot() +
  labs(
    title = "Distribution of Snowfall Values (> 0 and < 100 mm) by Year",
    x = "Year",
    y = "Snowfall (mm)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
*Comments:
0 is the most commonly observed value for snowfall. This actually makes sense because it usually snows during winter, which leads 0 value to be of highest frequency in dataset.

```{r}
#two-panel plot showing the average max temperature in Jan and Jul in each station across years
ny_noaa_clean <- ny_noaa_df %>%
  filter(month %in% c(1, 7)) %>% 
  select(id, year, month) %>% 
  mutate(mean_tmax = mean(ny_noaa_df$tmax, na.rm = TRUE))
```

#Problem 2

```{r}
#Import dataset accelerometer 
accelerometer <- read.csv("Problem2/nhanes_accel.csv")%>%
  janitor::clean_names()
View(accelerometer)
```

```{r}
#Import dataset demographic
demo <- read.csv("Problem2/nhanes_covar.csv",skip = 4,na = c("NA", "", "."))%>%
  janitor::clean_names()
view(demo)
```

```{r}
#Merge two datasets
participants <- left_join(accelerometer,demo, by = 'seqn')
view(participants)
```

```{r}
#data cleaning
participants_clean <- participants %>%
  relocate(sex,age,bmi,education,.before = min1)%>%
  filter(age >= 21) %>%
  drop_na(sex,age,bmi,education)%>%
  mutate(
    sex = factor(sex, levels = c(1, 2), labels = c("Male", "Female")),
    education = factor(education, levels = c(1, 2, 3), labels = c("Less than high school", "High school equivalent", "More than high school")),
    age = as.numeric(age),
    bmi = as.numeric(bmi))
```

```{r}
#table for men and women in each education category
sex_distribution <- participants_clean%>%
  group_by(education, sex) %>%
  summarize(count = n())

sex_distribution_table <- pivot_wider(
  sex_distribution,
  names_from = sex,
  values_from = count,
  values_fill = list(count = 0))

knitr::kable(sex_distribution)
```


```{r}
#age distribution in each education category
age_distribution <- participants_clean%>%
  group_by(education, sex, age)%>%
  select(education, sex, age)

ggplot(age_distribution, aes(x = age, fill = sex)) +
  geom_histogram(binwidth = 5, position = "dodge", alpha = 0.7) +
  facet_wrap(~ education) +
  labs(title = "Age Distribution by Sex in Each Education Category",
       x = "Age",
       y = "Count")
ggsave("age_distribution.png")
```

*Comments:
*For the reader-friendly table: the distribution of women and men is nearly equal in 'less than high school'. In 'high school equivalent', there are more males than females. In "more than high school', there are slightly more females than males.
*For age distribution plot: The distribution of age in males and females is nearly equal in 'less than high school'. The overall age of females is higher than the overall age of males and males have relatively lower median and mean. The distribution of age by sex is nearly the same in 'more than high school' while the median of age in females is lower than the the median of age in males.

```{r}
#Plot total activity against age in each education category
total_activity <- participants_clean%>%
  mutate(
    total_activity = rowSums(select(., starts_with("min")), na.rm = TRUE))%>%
  relocate(total_activity,.before = min1)%>%
  select(seqn,sex,age,education,total_activity)


ggplot(total_activity, aes(x = age,y = total_activity, color = sex)) +
  geom_smooth(se = FALSE)+
  geom_point(size = 2, position = position_dodge(width = 0.5), alpha = 0.7)+
  facet_wrap(~ education) +
  labs(title = "total activity by Sex in Each Education Category",
       x = "Age",
       y = "total_activity")
ggsave("total_activity.png")
```
*Comments: In each education category, there is a general decreasing trend for total activity as ages increases.In the "Less than high school" group, physical activity decreases with age for both men and women, with some fluctuations around ages 50-60, and individual variations are evident.

In the "High school equivalent" group, women generally have higher activity levels than men, particularly between ages 30-50. Activity decreases after age 40 for both sexes but increases before 40.

In the "More than high school" group, activity patterns are stable across ages, with women typically more active than men. Activity declines for both sexes after age 60.



```{r}
#24-hour activity time course for each education level and gender
mean_min_by_sex_education <- participants_clean %>%
  group_by(sex, education) %>%
  summarise(across(starts_with("min"), mean, na.rm = TRUE))

mean_min_long <- mean_min_by_sex_education %>%
  pivot_longer(
    cols = starts_with("min"), 
    names_to = "minute", 
    names_prefix = 'min',
    values_to = 'mean'
  )%>%
  mutate(minute = as.numeric(minute))

ggplot(mean_min_long, mapping=aes(x=minute, y=mean, color=sex)) + geom_line(size=0.5) + theme_light() + facet_grid(.~ education) +
  geom_smooth(size=0.7) + scale_x_continuous(limits=c(0, 1440)) +
  scale_y_continuous(limits=c(0,18), breaks=seq(0,18, by=3)) + xlab("24-hour activity time courses") + ylab("mean acceleorometers value per minute") +
  ggtitle("Scatterplot of  the mean acceleorometers readings per minute against 24-hour activity ") + theme(legend.position = "bottom", plot.title=element_text(hjust=.5))
ggsave("mean_min_long.png")
```
*Comments
The 24-hour activity time courses are similar in each education category with peak at 750 and lowest point at 250.
Less than high school: Both men and women have low activity levels in the early morning (around 250min) and peak in the afternoon followed by a steady decline until night.
High school equivalent: Both men and women also show low activity in the early morning, with peaks in the afternoon then decrease steadily into the night.
More than high school: The trend is similar to the "High school equivalent" group, but women show significantly higher activity levels than men during peak periods.

#Problem3


```{r}
#Import 4 datasets
Jan_2020 <- read_csv("Problem3/Jan 2020 Citi.csv",na = c("NA", "", ".","unknown"))%>%
  janitor::clean_names()%>%
  mutate(
    year = 2020, month = 'January')

Jan_2024 <- read_csv("Problem3/Jan 2024 Citi.csv",na = c("NA", "", ".","unknown"))%>%
  janitor::clean_names()%>%
  mutate(
    year = 2024, month = 'January')

July_2020 <- read_csv("Problem3/July 2020 Citi.csv",na = c("NA", "", ".","unknown"))%>%
  janitor::clean_names()%>%
  mutate(
    year = 2020, month = 'July')

July_2024 <- read_csv("Problem3/July 2024 Citi.csv",na = c("NA", "", ".","unknown"))%>%
  janitor::clean_names()%>%
  mutate(
    year = 2024, month = 'July')
```

```{r}
#merge 4 datasets
nyc_rider_df <- bind_rows(Jan_2020,July_2020,Jan_2024,July_2024)%>%
  relocate(year, month,.before = weekdays)
```

```{r}
nrow(nyc_rider_df)
summary(nyc_rider_df)
variable.names(nyc_rider_df)
```

*Description of dataset:
The cleaned nyc_rider dataset contains 99485 observations and 9 variables, containing ride information in January and July in 2020 and 2024. Key variables include "ride_id","rideable_type","year","month"           "weekdays","duration","start_station_name","end_station_name","member_casual" as indicator of member rider or casual rider status. The cleaning process include merging four corresponding datasets and cleaning column names.

```{r}
#produce table showing total number of riders in each year and month seprating casual rider and member

total_rider <- nyc_rider_df%>%
  mutate(year_month = str_c(year,month,sep = "_"))%>%
  select(ride_id,year_month,member_casual)%>%
  group_by(year_month,member_casual)%>%
  summarize(count = n())

total_rider_table <- pivot_wider(
  total_rider,
  names_from = year_month,
  values_from = count)

knitr::kable(total_rider_table)
  
```
*Comments: The dataset summarized the number of casual and member rides for two specific time points: January and July of 2020 and 2024.

Casual users: The number of casual rides increased significantly over time, from 984 in January 2020 to 10,894 in July 2024. There is also a noticeable seasonal difference, with higher counts in July compared to January in both years.

Member users: Member rides consistently outnumber casual rides, indicating potential benefits of being member. Their numbers also grew from 11,436 in January 2020 to 36,262 in July 2024. Similar to casual rides, there are higher counts in July than in January each year.

Both casual and member rides increased over the years, with a clear seasonal pattern showing higher activity in July. 


```{r}
#checking na in dataset July_2024
na_rows <- July_2024[!complete.cases(July_2024), ]

#Since several starting stations contain null value, I decided to drop na
July_2024_clean <- July_2024%>%
  drop_na(start_station_name)%>%
  select(ride_id,start_station_name)

#Create table showing 5 most popular starting stations for July 2024
July_2024_clean_table <- July_2024_clean%>%
  group_by(start_station_name)%>%
  summarize(count = n())%>%
  arrange(desc(count))
view(July_2024_clean_table)
```


```{r}
#Create a table including day of the week,month,year, median ride duration

median_duration <- nyc_rider_df %>%
  mutate(weekdays = factor(weekdays, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))) %>%
  select(year, month, weekdays, duration) %>%
  group_by(year, month, weekdays) %>%
  summarise(median_duration = median(duration, na.rm = TRUE))

ggplot(median_duration,aes(x = weekdays, y = median_duration, group = month,color = month))+
  geom_line(aes(color = month),alpha = .5)+
  facet_grid(.~year)+
  labs(title = "Median Ride Duration by Weekday, Month, and Year",
       x = "Day of the Week",
       y = "Median Ride Duration (minutes)",
       color = "Month")+
   theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(legend.position = "bottom")
ggsave("median_duration.png")
```
*Comments:
The median ride duration is much higher in 2020 compared to 2024.
The median ride duration in July tends to be higher than median ride duration in January.
The median ride duration tends to be higher in weekends comparing to weekdays.

```{r}
#Figure that shows the impact of month,membership status, bike type on distribution of duration in 2024

bikes_2024 <- nyc_rider_df%>%
  filter(year == 2024)%>%
  group_by(rideable_type,month,member_casual)

bikes_2024_plot <- ggplot(bikes_2024,aes(x = duration,fill = factor(member_casual))) +
  geom_histogram(binwidth =10, position = "dodge", alpha = 0.9) +
  facet_grid(month ~ rideable_type)+
     labs(
       title = "Histogram of Ride Duration by Member Status, Month, and Bike Type 2024",
       x = "Ride Duration (minutes)",
       y = "Count",
       fill = "Member_casual")

print(bikes_2024_plot)
ggsave("bikes_2024_plot.png")
```
*Comments:
The distribution of electric bikes is higher for shorter rides compared to classic bikes. 
Member tend to have higher ride duration compared to casual rides.
July has significantly more rides compared to January.
Even though casual rides have lower ride duration, it is more spread compared to member rides.

